Improving Haskell Theory Exploration

### 2015-12-01 TransferReport/background.tex
Fleshed out feature extraction

As an example, we can consider a simple signature for Peano numerals, addition and multiplication:

\begin{align*}
  \Sigma_{\mathbb{N}} = \{ & 0     && : \mathbb{N}, \\
                        & succ  && : \mathbb{N} \rightarrow \mathbb{N}, \\
                        & plus  && : \mathbb{N} \rightarrow \mathbb{N} \rightarrow \mathbb{N}, \\
                        & times && : \mathbb{N} \rightarrow \mathbb{N} \rightarrow \mathbb{N}\}
\end{align*}

Together with a set of variables, say $V_{\mathbb{N}} = \{x, y, z\}$, \qspec{}'s enumeration will resemble the following:

\begin{align*}
  terms_{\mathbb{N}} = [& 0,\ succ,\ plus,\ times,\ x,\ y,\ z,\ succ(0),\ succ(x),\ succ(y), \\
                     & succ(z),\ plus(0,\blank),\ plus(x, \blank),\ \dots ]
\end{align*}

Notice that our functions are curried, where $f(x,\blank)$ indicates \emph{partial application} of $f$ to $x$. This is required as the construction of $terms$ applies functions to one argument at a time.

\begin{figure}
  % To reproduce, run 'quickSpec nat' in haskell_example/src/QuickSpecExample.hs
  \begin{align*}
                        plus(x, y) &= plus(y, x)            \\
                        plus(x, 0) &= x                     \\
               plus(x, plus(y, z)) &= plus(y, plus(x, z))   \\
                       times(x, y) &= times(y, x)           \\
                       times(x, 0) &= 0                     \\
             times(x, times(y, z)) &= times(y, times(x, z)) \\
                  plus(x, succ(y)) &= succ(plus(x, y))      \\
                 times(x, succ(y)) &= plus(x, times(x, y))  \\
              times(x, plus(y, y)) &= times(y, plus(x, x))  \\
    plus(times(x, y), times(x, z)) &= times(x, plus(y, z))
  \end{align*}
  \caption{Equations conjectured by \qspec{} for Peano numerals, addition and multiplication; after simplification.}
  \label{fig:qspecresult}
\end{figure}

These terms will be grouped into three classes, one each for $\mathbb{N}$, $\mathbb{N} \rightarrow \mathbb{N}$ and $\mathbb{N} \rightarrow \mathbb{N} \rightarrow \mathbb{N}$. As the variables $x$, $y$ and $z$ are instantiated to various randomly-generated numbers, these equivalence classes will be divided, until eventually the equations in figure \ref{fig:qspecresult} are conjectured.

@@

\subsection{Feature Extraction}
\label{sec:featureextraction}

One of our main contributions is a \emph{feature extraction} algorithm for Core expressions. Feature extraction is the conversion of ``raw'' input data, such as audio, images or (in our case) expressions, into a form more suited to machine learning algorithms. By pre-processing our data in this way, we can re-use the same ``off-the-shelf'' machine learning algorithms in a variety of domains.

We use a standard representation of features known as a \emph{feature vector}: a sequence of numbers $(x_1, \dots, x_n)$ where $x_i \in \mathbb{R}$. \footnote{In fact, we only use $\mathbb{R}$ for notational and analytic convenience. Practical implementations will use an approximate format such as IEEE 754 floating point numbers.} For learning purposes this has two important advantages over our raw expressions:

\begin{itemize}
  \item All of our feature vectors will be the same size, i.e. they will all have length (or \emph{dimension}) $n$. This makes element-wise comparison between arbitrary feature vectors $x$ and $y$ trivial, as each element $x_i$ will have a corresponding element $y_i$. For raw expressions this is not so straightforward, as their nesting may give rise to very different shapes.
  \item Unlike our discrete expressions, we can continuously transform one feature vector into another. This enables many powerful ML algorithms to be used, such as those based on \emph{gradient descent} or, in our case, taking the arithmetic mean.
\end{itemize}

Our approach to scaling up these Haskell theory exploration tools takes inspiration from two sources. The first is premise selection, which makes expensive algorithms used in theorem proving more practical by limiting the size of their inputs. We describe this approach in more details in \S \ref{sec:relevance}. Premise selection is a practical tool which has seen use in practical software, such as the \emph{Sledgehammer} component of the Isabelle/HOL theorem prover.

Despite the idea's promise, we cannot simply invoke existing premise selection algorithms in our theory exploration setting. The reason is that theory exploration has no distinct \emph{goal} \iffalse TODO: implies that premise selection has a goal; rephrase: premise selection is given a concrete formula (called the ``goal'') to compare terms against \fi to compare expressions against. Instead, we are interested in relationships between \emph{any} terms generated from a signature, and hence we must consider the relevance of \emph{all terms} to \emph{all other terms}. A natural fit for such problems is \emph{clustering}, which attempts to group similar inputs together in an unsupervised way. See \S \ref{sec:clustering} for more information.

Our hypothesis is that clustering methods, as found in systems like ML4PG and ACL2(ml), can be used as relevance filters to break up large signatures into smaller ones more amenable to brute force enumeration.

\iffalse TODO: I would play up the clustering intelligence. Clustering is not just about breaking up, it is about discovering significant patterns in data. By forgetting this, you make motivation for your work sound too ``small'' \fi

@@

\subsection{Clustering}
\label{sec:clustering}
% TODO

Clustering is an unsupervised machine learning task for grouping $n$ data points using a similarity metric. There are many variations on this theme, but in our case we will fix the number of clusters at \iffalse TODO: use ceiling brackets \fi $k = \sqrt{n}$ the data points will be $d$-dimensional feature vectors of rational numbers, the similarity metric is euclidean distance, denoted $d$. and clustering will be performed by the $k-means$ algorithm. We choose this setup as it is well studied and supported by off-the-shelf tools, and is used in the ML4PG and ACL2(ml) systems which served as inspiration for this work.

We will use $S$ to denote our set of $n$ data points and $k$ as the number of clusters.

\subsubsection{k-Means}

The k-means algorithm works, as the name suggests, by calculating the \emph{mean value} of each cluster's elements, which we will denote as $m_1$ to $m_k$. Since k-means is iterative, we will use superscript notation to denote time steps, such as $m_i^t$ for the value of $m_i$ at step $t$.

Before k-means starts, we must choose \emph{seed} values for $m_1^0$ to $m_k^0$. Many methods have been proposed for choosing these values, e.g. see CITE. For simplicity, we will assume they are chosen at random from our data set. Subsequent values, i.e. $m_i^t$ for $t > 0$, are simply the mean values of the corresponding cluster $C_i^t$:

TODO: We need to handle the vectors element-wise, but we've already used subscript notation!

\begin{equation}
  m_i^t =
\end{equation}

>>>>

As an example, we can consider a simple signature containing the expressions from figure \ref{fig:haskellexample}:

\begin{align*}
  \Sigma_{\texttt{Nat}} = \{\texttt{Z}, \texttt{S}, \texttt{plus}, \texttt{mult}, \texttt{odd}, \texttt{even}\}
\end{align*}

Together with a set of variables, say $V_{\texttt{Nat}} = \{a, b, c\}$, \qspec{}'s enumeration will resemble the following:

\begin{align*}
  terms_{\texttt{Nat}} = [& \texttt{Z},\ \texttt{S},\ \texttt{plus},\ \texttt{mult},\ \texttt{odd},\ \texttt{even},\ a,\ b,\ c,\ \texttt{S Z},\ \texttt{S}\ a,\ \texttt{S}\ b, \\
                     & \texttt{S}\ c,\ \texttt{plus Z},\ \texttt{plus}\ a,\ \dots ]
\end{align*}

Notice that Haskell curries functions, so the binary functions \hs{plus} and \hs{mult} can be treated as unary functions which return unary functions. This is required as the construction of $terms$ applies functions to one argument at a time.

\begin{figure}
  % To reproduce, run 'quickSpec nat' in haskell_example/src/QuickSpecExample.hs
  \begin{haskell}
                  plus a b = plus b a
                  plus a Z = a
         plus a (plus b c) = plus b (plus a c)
                  mult a b = mult b a
                  mult a Z = Z
         mult a (mult b c) = mult b (mult a c)
              plus a (S b) = S (plus a b)
              mult a (S b) = plus a (mult a b)
         mult a (plus b b) = mult b (plus a a)
                 odd (S a) = even a
            odd (plus a a) = odd Z
           odd (times a a) = odd a
                even (S a) = odd a
           even (plus a a) = even Z
          even (times a a) = even a
plus (mult a b) (mult a c) = mult a (plus b c)
  \end{haskell}
  \caption{Equations conjectured by \qspec{} for the functions in figure \ref{fig:haskellexample}; after simplification.}
  \label{fig:qspecresult}
\end{figure}

These terms will be grouped into five classes, one each for \hs{Nat}, \hs{Nat -> Nat} and \hs{Nat -> Nat -> Nat}, \hs{Nat -> Bool} and \hs{Bool}. As the variables $a$, $b$ and $c$ are instantiated to various randomly-generated numbers, these equivalence classes will be divided, until eventually the equations in figure \ref{fig:qspecresult} are conjectured.

@@

\subsection{Clustering}
\label{sec:clustering}

Our approach to scaling up these Haskell theory exploration tools takes inspiration from two sources. The first is premise selection, which makes expensive algorithms used in theorem proving more practical by limiting the size of their inputs. We describe this approach in more details in \S \ref{sec:relevance}. Premise selection is a practical tool which has existing applications in software, such as the \emph{Sledgehammer} component of the Isabelle/HOL theorem prover.

Despite the idea's promise, we cannot simply invoke existing premise selection algorithms in our theory exploration setting. The reason is that premise selection requires a distinguished expression to compare everything against; in practice this is the current goal of a theorem prover. Theory exploration does not have such a distinguished expression; instead, we are interested in relationships between \emph{any} terms generated from a signature, and hence we must consider the relevance of \emph{all terms} to \emph{all other terms}.

A natural fit for this task is \emph{clustering}, which attempts to group similar inputs together in an unsupervised way. Based on their success in discovering relationships and patterns between expressions in Coq and ACL2 (in the ML4PG and ACL2(ml) tools respectively), we hypothesise that clustering methods can fulfil the role of relevance filters for theory exploration: intelligently breaking up large signatures into smaller ones more amenable to brute force enumeration, such that related expressions are explored together.

\subsubsection{Feature Extraction}
\label{sec:featureextraction}

Before describing clustering in detail, we must introduce the idea of \emph{feature extraction}. This is the conversion of ``raw'' input data, such as audio, images or (in our case) Core expressions, into a form more suited to machine learning algorithms. By pre-processing our data in this way, we can re-use the same ``off-the-shelf'' machine learning algorithms in a variety of domains.

We use a standard representation of features as a \emph{feature vector} of numbers $(x_1, \dots, x_p)$ where $x_i \in \mathbb{R}$. \footnote{In fact, we only use $\mathbb{R}$ for notational and analytic convenience. Practical implementations will use an approximate format such as IEEE 754 floating point numbers.} For learning purposes this has two important advantages over our raw expressions:

\begin{itemize}
  \item All of our feature vectors will be the same size, i.e. they will all have length (or \emph{dimension}) $p$. This makes element-wise comparison between arbitrary feature vectors $\mathbf{x}$ and $\mathbf{y}$ trivial: compare each element $x_i$ with $y_i$ for $1 \leq i \leq p$. For expressions would not be so straightforward, as their nesting may give rise to very different shapes.
  \item Unlike our expressions, which are discrete, we can continuously transform one feature vector into another. This enables many powerful ML algorithms to be used, such as those based on \emph{gradient descent} or, in our case, arithmetic means.
\end{itemize}

\subsubsection{K-Means}
\label{sec:kmeans}

Clustering is an unsupervised machine learning task for grouping $n$ data points using a similarity metric. There are many variations on this theme, but in our case we make the following choices:

\begin{itemize}
  \item We fix the number of clusters at $k = \lceil \sqrt{n} \rceil$.
  \item Data points will be $p$-dimensional feature vectors, as defined above.
  \item We will use euclidean distance (denoted $d$) as our similarity metric.
  \item Clustering will be performed by the $k-means$ algorithm.
\end{itemize}

This is a standard setup, supported by off-the-shelf tools. In particular, it is similar to that used by the ML4PG and ACL2(ml) systems which served as inspiration for this work, which makes our results more easily comparable.

The k-means algorithm works, as the name suggests, by calculating the \emph{mean value} of each cluster's elements. Since k-means is iterative, we will use function notation to denote time steps, so $x(t)$ denotes the value of $x$ at time $t$. We denote the clusters as $C^1$ to $C^k$, their mean values as $\mathbf{m}^1$ to $\mathbf{m}^k$. Subscripts are vector elements, so $m^i_j(t)$ is the $i$th element of the $j$th mean, at time $t$.

Before k-means starts, we must choose \emph{seed} values for $\mathbf{m}^1(0)$ to $\mathbf{m}^k(0)$. Many methods have been proposed for choosing these values, e.g. see \citep{arthur2007k}. For simplicity, we will assume they are chosen at random from our data set. Subsequent values, i.e. $\mathbf{m}^i(t)$ for $t > 0$, are simply the mean values of the corresponding cluster $C^i(t)$:

\begin{equation}
  m^i_j(t) = \frac{\sum_{x \in C^i(t)} x_j}{|C^i(t)|}
\end{equation}

@@

### 2015-12-01 TransferReport/contributions.tex
Fleshed out feature extraction

The purpose of recurrent clustering, as with other clustering algorithms, is to identify similarities between a set of inputs. In our case, we want compare the abstract syntax trees of Core expressions, which presents a difficulty for clustering: many leaves are \emph{references}, containing identifiers for other expressions. For example, consider the following expressions which are identical except for an identifier:

@@

Two possible methods for comparison are to ignore identifiers completely, or to compare them for equality. Neither is completely satisfactory, as the former will \emph{over-estimate} similarity (considering $X$, $Y$ and $Z$ to be the same), whilst the latter will \emph{under-estimate} similarity (considering $X$, $Y$ and $Z$ to be equidistant).

>>>>

The purpose of recurrent clustering, as with other clustering algorithms, is to identify similarities between a set of inputs. In our case, we want to compare Core expressions, which presents a difficulty for clustering: many leaves are \emph{references}, containing identifiers for other expressions. For example, consider the following expressions which are identical except for the global identifiers (which refer to the expressions in figure \ref{fig:coreexample}):

@@

We saw in section Two possible methods for comparison are to ignore identifiers completely, or to compare them for equality. Neither is completely satisfactory, as the former will \emph{over-estimate} similarity (considering $X$, $Y$ and $Z$ to be the same), whilst the latter will \emph{under-estimate} similarity (considering $X$, $Y$ and $Z$ to be equidistant).
